---
title: "Getting Started with PeakRankR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PeakRankR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # bigwig files not available in vignette build
)
```

## Overview

**PeakRankR** prioritises genomic enhancer peaks across cell groups for
experimental cloning. Given:

- a table of peak coordinates with group labels (magnitude is optional — auto-computed from bigwig if omitted), and
- a table pointing to bigwig files per group,

it scores each peak on **specificity**, **sensitivity**, and **magnitude**,
then returns the peaks ranked within each group.

### Scoring formula

$$
\text{Score} = W_{\text{spec}} \times \text{Specificity}
             + W_{\text{sens}} \times \text{Sensitivity}
             + W_{\text{mag}}  \times \text{Magnitude}
$$

All three components are min-max normalised to [0, 1] before weighting.

---

## Installation

```r
# 1. System dependency — bedtools
#    macOS:   brew install bedtools
#    Ubuntu:  sudo apt-get install bedtools
#    Conda:   conda install -c bioconda bedtools

# 2. Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("rtracklayer", "GenomicRanges", "IRanges", "S4Vectors"))

# 3. PeakRankR
devtools::install_github("AllenInstitute/PeakRankR", dependencies = TRUE)
```

---

## Quick start

```{r load}
library(PeakRankR)
```

### Step 1 — Verify bedtools

```{r check_bedtools}
# Auto-detect on PATH, or set: options(bedtools.path = "/usr/local/bin/")
check_bedtools()
```

### Step 2 — Load input files

The **peak TSV** must have at minimum: chromosome, start, end, and a group column.
Magnitude is optional — if not provided it is computed automatically from the bigwig signal. Column names are flexible — pass the actual names via function
arguments.

```{r load_data}
# Built-in example files
tsv_path <- system.file("extdata", "test_file.tsv", package = "PeakRankR")
bw_path  <- system.file("extdata", "bw_table.txt",  package = "PeakRankR")

peaks    <- read.table(tsv_path, header = TRUE, sep = "\t")
bw_table <- read.table(bw_path,  header = TRUE, sep = "\t")

head(peaks)
#>    chr    start      end cell.population
#> 1 chr6 31131000 31131500    Astrocytes-1
#> 2 chr6 31126700 31127200    Astrocytes-1
```

The **bigwig table** must have exactly two columns: `file_path` and `sample_id`.
The `sample_id` values must match values in your group column.

```{r bw_table}
# Resolve the bundled bigwig filenames to full installed paths
extdata_path <- system.file("extdata", package = "PeakRankR")
bw_table$file_path <- file.path(extdata_path, bw_table$file_path)

head(bw_table)
#>                                                        file_path sample_id
#> 1 /path/to/R/library/PeakRankR/extdata/Astrocytes-1.bw Astrocytes-1
#> 2 /path/to/R/library/PeakRankR/extdata/Astrocytes-2.bw Astrocytes-2
```

### Step 3 — Run ranking

```{r rank}
ranked <- Peak_RankR(
  tsv_file_df          = peaks,
  group_by_column_name = "cell.population",
  background_group     = unique(peaks$cell.population),
  bw_table             = bw_table,
  rank_sum             = TRUE,
  weights              = c(1, 1, 1)
)

# View top-ranked peaks per group
ranked[order(ranked$PeakRankR_rank), ]
```

---

## Custom column names

If your data uses different column names, specify them:

```{r custom_cols}
ranked <- Peak_RankR(
  tsv_file_df          = my_peaks,
  group_by_column_name = "subclass",   # your group column
  bw_table             = bw_table,
  chr_col              = "chrom",
  start_col            = "chromStart",
  end_col              = "chromEnd",
  magnitude_col        = "fold_change"
)
```

---

## Adjusting weights

To prioritise specificity over the other features:

```{r weights}
ranked_specific <- Peak_RankR(
  tsv_file_df          = peaks,
  group_by_column_name = "cell.population",
  bw_table             = bw_table,
  weights              = c(3, 1, 1)   # specificity weight tripled
)
```

---

## Output columns

`Peak_RankR()` returns the input data frame with five new columns appended:

| Column | Description |
|---|---|
| `specificity_score` | Normalised ratio of target signal to background |
| `sensitivity_score` | 1 - (groups with signal / total groups) — higher means fewer groups share this peak |
| `magnitude_score` | Normalised mean signal level |
| `PeakRankR_score` | Weighted composite score |
| `PeakRankR_rank` | Rank within the group (1 = best candidate) |

---

## Session info

```{r session}
sessionInfo()
```
